---
title: "Setup: First Steps with R Studio"
subtitle: "Do this before Day 1"
execute:
  eval: false
---

Welcome to this R Studio tutorial. It's intended to support the
in-person walkthrough and is a bit more detailed, with exercises
included along the way.

## Download R and R Studio

To use R Studio, first download R, then R Studio — in that order. R is
the programming language; R Studio is the interface. Get R directly from [CRAN](https://cran.r-project.org/) and R Studio from here [posit.co/download/rstudio-desktop](https://posit.co/download/rstudio-desktop/). Make sure the R version matches your operating system and, on a Mac, your processor. After
downloading, the installer should start automatically.

## R Studio interface & Markdown

On the right side of the screen is the **console** — the window where
commands are executed. Executed code and any error messages appear
here, but the commands themselves are not saved. On the upper left is
the **Environment** tab — your workspace, listing all relevant
objects, variables, and data sets currently in use. Below that there
are several more tabs; the **Help** tab is particularly useful, giving
access to function documentation.

## What are packages?

Packages are collections of commands, which must be loaded into R
Studio before they can be used, the same way you have to be taught
something before you can apply it.

Installing a package (name in quotation marks, since it's a string):

```{r}
# install.packages("pacman")
```

Packages only need to be installed once. Before an already-installed
package can be used, it needs to be **loaded**, this time without
quotation marks:

```{r}
library(pacman)
```

Loading packages one by one every session is tedious. The **pacman**
package solves this: `p_load()` installs *and* loads packages in one
call.

```{r}
pacman::p_load("haven", "tidyverse", "rmarkdown")
```

`haven` lets us import files saved in Stata format (`.dta`).

Datasets also need to be loaded into R Studio, from a file saved on
your computer (or a URL). Every file has a **path** — its address —
which we need to specify in order to read it in.

## Creating objects

R is an object-based language: we store information in objects, which
are containers for values. This could be a single variable (e.g.
"age") or a whole dataset (rows = observations/cases, columns =
variables).

Below, we create an object called `fruit` (a vector) using the
assignment arrow `<-`. `c()` ("combine") lists the values it should
contain.

```{r}
fruit <- c("apple", "pear", "grapes")

fruit[2]

price <- c(2, 1, 4)
```

R is **case-sensitive** — `"h"` and `"H"` are different characters. It
helps to write everything in lowercase to avoid this source of error.

## Object classes

Different objects hold different classes of values, which determine
what operations are valid on them. A character object supports only
very limited mathematical operations, taking the square root of
`"Hello World"` doesn't make sense, and R will tell you so.

```{r}
str    <- "Hello World"
num    <- 2.3
num_2  <- 4
bool   <- TRUE
bool_2 <- FALSE
```

Check an object's class with `class()`:

```{r}
class(str)
class(num)
class(num_2)
class(bool)
class(bool_2)
```

We can also calculate with objects, often not just a single value,
but a whole vector:

```{r}
# sqrt(str)   # would error — str is a character, not numeric
num + num_2
```

## Creating a dataset

Beyond single vectors, we can build small datasets — e.g. combining
the `fruit` and `price` vectors with `cbind()`:

```{r}
basket <- cbind(fruit, price)
basket <- as.data.frame(basket)
View(basket)
```

With this mini dataset we can already run some calculations:

```{r}
mean(price)     # arithmetic mean
median(price)   # median
min(price)      # smallest value in the column
max(price)      # largest value in the column
sum(price)      # sum of the column
summary(price)  # all of the above, at once
```

### Exercise

1. Create a dataset from any two variables. Three cases is plenty. One
   variable should hold numeric values, the other characters.

## Importing a dataset

**Working directory** — the path where your dataset lives. Once set,
you only need to specify the file name, not its full path. Set it with
`setwd()`; check the current one with `getwd()`. An incorrect working
directory is a common reason a dataset fails to load.

**Finding a file's path** — on a Mac, click the file in Finder, then
right-click (or Control-click) and choose "Copy … as Pathname." On
Windows, right-click the file → Properties → "Location" gives the
folder, and you still need to append the file name and extension.
Windows paths use double backslashes (`\\`) instead of a single slash
(`/`).

We'll use the **Titanic** dataset (887 passengers) for the exercises
below. It's in Stata's `.dta` format, readable in R via the `haven`
package's `read_dta()` (a plain CSV would use `read.csv()` instead).

[⬇ Download `Titanic.dta` (92 KB)](../data/Titanic.dta){.btn .btn-outline-primary role="button" download="Titanic.dta"}

```{r}
#| label: import-data
library(haven)
df <- read_dta("../data/Titanic.dta")
```

Check the Environment pane (top right) to confirm the dataset loaded —
it'll show the number of observations and variables.

```{r}
View(df)
```

If variable names aren't self-explanatory, check the dataset's
**codebook** for more detailed descriptions. Always review a dataset
before using it.

### Overview of variables

```{r}
ncol(df)  # number of columns
nrow(df)  # number of rows
```

```{r}
names(df)  # variable names
```

Access a specific variable with `$`:

```{r}
summary(df$age)
```

## Creating tables

`table()` builds a frequency table; `prop.table()` converts it to
proportions. To interpret these you need to know how the variables are
coded.

```{r}
gender_distribution <- table(df$sex)
gender_distribution

proptable <- prop.table(gender_distribution)
round(proptable, digits = 2)
```

```{r}
survived_distribution <- table(df$survived)
survived_distribution

proptable <- prop.table(survived_distribution)
round(proptable, digits = 2)
```

**Cross-tabulation** — same `table()` function, just add a second
variable after the comma:

```{r}
tab_2 <- table(df$sex, df$survived)  # sex x survived
tab_2
```

The result isn't very readable without knowing the coding — here,
`survived`: 0 = died, 1 = survived. Add labels with `dimnames()`:

```{r}
dimnames(tab_2) <- list(sex = c("female", "male"), survived = c("died", "survived"))
tab_2
```

("Women and children first": 74% of women survived, vs. 19% of men.)

```{r}
proptable <- prop.table(tab_2)
round(proptable, digits = 2)

# proptable_cols <- prop.table(tab_2, margin = 2)  # column-wise proportions
```

### Exercises

**Importing data**

2. Load the Titanic dataset.

**Describing data**

3. What object class does the variable `pclass` (passenger class)
   have?
4. How many women and men were on the Titanic?

**Descriptive results**

5. What was the average ticket price (median and mean; variable
   `fare`)?
6. What was the average age of the Titanic passengers?
7. How many children did Mr. Anders Johan Andersson have on board?

## Visualizations

`hist()` gives a quick overview of a variable's distribution:

```{r}
hist(df$age)
```

For more flexible graphics, use **ggplot2** — see this
[cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/data-visualization.pdf).
A boxplot:

```{r}
# Preprocessing before plotting
df$sex <- as.factor(df$sex)
df <- df[!is.na(df$sex), ]
df <- df[!is.na(df$age), ]

df %>%
  ggplot(aes(x = sex, y = age, group = sex)) +
  geom_boxplot()
```

### Exercises

8. Visualize the distribution of a variable.
9. Visualize a variable a different way.

## Piping

Data outside of seminars is rarely clean — it needs preparation. For
**data wrangling**, chain multiple commands together with piping.
Assign to a new object with `<-`, specify the dataset, then chain
tidyverse commands with `%>%` (Mac: Shift+Cmd+M, Windows: Ctrl+Shift+M).
Pressing Enter after a pipe automatically indents the next line — no
need to repeat the dataset name, and the last line doesn't need a
trailing pipe.

```{r}
df_clean <- df %>%
  select(sex, age, fare, survived) %>%  # pick specific columns
  filter(sex == "female") %>%            # pick specific rows
  rename(gender = "sex")                 # rename a variable

# na.omit(df_clean$fare)  # drop missing values
```

### What's the difference between `=` and `==`?

Two equals signs (`==`) make a comparison and return `TRUE`/`FALSE`. A
single equals sign is an assignment operator.

### Exercises

**Data preprocessing**

10. Rename the variable `sex` to `gender`.
11. Create a sub-dataset with at least three variables, containing
    only passengers who survived the Titanic.
12. Remove all missing values from the dataset.

## If-else statements

Most data wrangling can be done with tidyverse functions, but
sometimes it's easier to think in if-then conditions.

We want a new variable that's `"low"` when the fare is at or below the
median, and `"high"` otherwise:

```{r}
df_clean$fare_category <- ifelse(df_clean$fare <= median(df$fare), "low", "high")
View(df_clean)
```

## Getting help with error messages

Error messages are completely normal in R — the cause isn't always
obvious, and it helps to look for support:

- The **Help** tab (bottom right) has documentation for specific
  functions and commands.
- [RPubs](https://rpubs.com/) and
  [Stack Overflow](https://stackoverflow.com/) are great community
  resources — code can often be copied and adapted.
- An LLM (e.g. ChatGPT) can help debug faulty code or explain what a
  snippet does, so you build a solid understanding rather than just
  copy-pasting.
